---
title: "Bayesian Inference for Discrete Random Variables"
subtitle: "Statistical Inference II"
author:
"Mine Dogucu"
"Noah Johnson"
date: "April 17, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(install.load)
install_load('tidyverse')
```

## Conditional Probability

$P(A|B) = \frac{P(A \cap B)}{P(B)} \implies P(A \cap B) = P(A|B) * P(B)$

$P(B|A) = \frac{P(A \cap B)}{P(A)} \implies P(A \cap B) = P(B|A) * P(A)$

## Bayes' Theorem

## Bayesian Universe

## Bayesian Inference by Tables

There is an urn containing a total of 5 balls, some of which are red and the rest of which are green. We don't know how many of the balls are red. Let $\theta$ be the number of red balls in the urn.

### Bayesian universe of the experiment

### Joint and marginal probability distributions

### Finding the posterior probability of $\theta | (Y = 1)$

### The posterior probability distribution after second red observation (no replacement).

## Consider an exam with 4 multiple-choice questions. Each question has 3 options. A student who hasn't studied guesses at each question with a probability of 1/3 of getting it right. Students who have studied might have probabilities 2/3 or 0.95 of getting questions right. Make a table of likelihoods. You may want to use R for this question.

```{r}
t <- tibble(theta = c(1/3, 2/3, 0.95), prior = rep(1/3, 3))

t$likelihood0 <- dbinom(0, 4, t$theta)
t$likelihood1 <- dbinom(1, 4, t$theta)
t$likelihood2 <- dbinom(2, 4, t$theta)
t$likelihood3 <- dbinom(3, 4, t$theta)
t$likelihood4 <- dbinom(4, 4, t$theta)

t

```

### You have observed that a student has responded to two questions correctly. What can you conclude about whether they have studied or not?

```{r}
ptl <- t$prior * t$likelihood2
marginal <- sum(ptl)

posterior <- ptl / marginal
posterior
```

We can't tell whether the student was just guessing or is a medium-strength student (theta = 2/3). But it's unlikely that they are an exceptional student (theta = 0.95).

### In another exam, you observe another student who has answered 50 out of 100 questions correctly. What can you conclude about whether they have studied or not?

```{r}
lik50 <- dbinom(50, 100, t$theta)
lik50
ptl <- t$prior * t$likelihood2
marginal <- sum(ptl)

posterior <- ptl / marginal
posterior
```

